---
title: "BRENDA & GEO platform"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "02/27/2024"
vignette: >
  %\VignetteIndexEntry{BRENDA & GEO platform}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<font color = "#000000">

```{r}
devtools::load_all(".")
```

We use BRENDA tissue ontology to indicate the tissue or cell-line of a signature, and GEO platform accession id for platform records.  
BRENDA tissue ontology version: **10/2021**. The OBO file was downloaded at https://www.brenda-enzymes.org/ontology.php.  
GEO platform IDs are manually downloaded at https://www.ncbi.nlm.nih.gov/geo/browse/?view=platforms. 25,827 platforms were downloaded as of 2024/02.  
If you are not sure which BRENDA tissue ontology or GEO platform accession id to use, you can search for them as shown below.  

## Search for a BRENDA tissue ontology 
Multiple search terms are seperated by space " ". \
Set `contain_all = TRUE` to show results include all search terms. \
Set `contain_all = FALSE` to show results include any of the search terms. \
```{r}
BRENDASearch("HEK 293 T", contain_all = TRUE)
```

## Search for a GEO platform
```{r}
GEOPlatformSearch("Illumina HiSeq 4000", organism = "Homo Sapiens")
```

Show only accession id:
```{r}
GEOPlatformSearch("Illumina HiSeq 4000", organism = "Homo Sapiens", accession_only = TRUE)
```

Search for multiple terms: 
```{r}
GEOPlatformSearch(c("Drosophila melanogaster", "Illumina HiSeq 4000"), organism = "Homo Sapiens", contain_all = TRUE)
```
Set `contain_all = TRUE` to show results include all search terms. \
Set `contain_all = FALSE` to show results include any of the search terms. \

</font>
--
