---
title: "BRENDA & GEO platform"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "03/31/2021"
vignette: >
  %\VignetteIndexEntry{BRENDA & GEO platform}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<font color = "#000000">

We use BRENDA tissue ontology to indicate the tissue or cell-line of a signature, and GEO platform accession id for platform records.  
The record was last updated **09/2020**, so some newest ontology and platform may not be included.  
If you are not sure which BRENDA tissue ontology or GEO platform accession id to use, you can search for them as shown below.  

## Search for a BRENDA tissue ontology
```{r}
devtools::load_all(".")
BRENDASearch("MDA")
```

## Search for a GEO platform
```{r}
GEOPlatformSearch("Illumina HiSeq 4000", species = "Homo Sapiens")
```

Show only accession id:
```{r}
GEOPlatformSearch("Illumina HiSeq 4000", species = "Homo Sapiens", accession_only = TRUE)
```

Search for multiple terms: 
```{r}
GEOPlatformSearch(c("Drosophila melanogaster", "Illumina HiSeq 4000"), species = "Homo Sapiens", contain_all = TRUE)
```
Set `contain_all = T` to show results include all search terms. Set `contain_all = FALSE` to show results include any of the search terms. 

</font>
--
