---
title: "Comparing OmicSignature Lists"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing OmicSignature Lists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This vignette outlines the workflow for comparing `OmicSignature` objects and
visualizing similarity among signatures. It is prepared for later publication
with the package website and uses small illustrative snippets rather than
bundled example signature files.

## Compare one list of signatures

Use `compare_omic_signatures()` with a named list of `OmicSignature` objects to
compare every signature against every other signature in the list. For
bi-directional signatures, the function compares the first factor level against
the first factor level and the second factor level against the second factor
level.

```{r}
library(OmicSignature)

data(compare_signatures_example)
signature_list <- compare_signatures_example

overlap_res <- compare_omic_signatures(
  sig_list1 = signature_list,
  method = "overlap",
  score_cutoff = log2(1.25),
  adj_p_cutoff = 0.05,
  min_features = 25,
  max_feature = 2000
)
## show the label order
overlap_res$label_order

## list the comparisons being performed
names(overlap_res$comparisons)

## show the comparison results for level 1 (jaccard and counts)
overlap_res$comparisons$level1_vs_level1$jaccard

## in the counts matrix, you can see the signature sizes in the diagonal
overlap_res$comparisons$level1_vs_level1$counts
```

The overlap method returns three matrices for each compared factor level:

* `jaccard`: Jaccard similarity between retained feature sets.
* `pvalue`: Fisher exact test p-values for overlap enrichment.
* `counts`: an integer matrix with the same dimensions as `jaccard`/`pvalue`.
  Entry `[i, j]` is the overlap size between signature `i` and signature `j`;
  for self-comparisons, the diagonal is therefore each signature's own
  retained feature-set size.

## Compare two lists of signatures

Pass both `sig_list1` and `sig_list2` to compare signatures across collections.
This is useful when signatures come from different studies, cohorts, platforms,
or perturbation screens.

```{r}
reference_signatures <- compare_signatures_example[1:2]
query_signatures <- compare_signatures_example[3:4]

cross_res <- compare_omic_signatures(
  sig_list1 = query_signatures,
  sig_list2 = reference_signatures,
  method = "overlap",
  score_cutoff = log2(1.25),
  adj_p_cutoff = 0.05,
  min_features = 25,
  max_feature = 500
)

cross_res$comparisons$level1_vs_level1$jaccard
```

When `background` is not supplied, the feature universe is inferred from all
available signature and differential-expression tables in both inputs. Supplying
a measured or assay-specific background is recommended when it is available.

## Control label pairing

By default, level ordering follows the factor levels in each signature's
`group_label` column. Use `label_pairing` and `label_pairing2` when signatures
need explicit matching.

```{r}
data(compare_label_pairing_example)

## let's see the order of the signatures' group_label's
compare_label_pairing_example |>
  sapply(\(x) levels(x$difexp$group_label)) |> t() |>
  as.data.frame() |> setNames(c("level1", "level2"))

## compare signatures and provide the desired label pairing
paired_res <- compare_omic_signatures(
  sig_list1 = compare_label_pairing_example,
  method = "overlap",
  label_pairing = list(
    signature_a = c("treated", "control"),
    signature_b = c("up", "down"),
    signature_c = c("resistant", "sensitive")
  )
)
## level1 vs. level1
paired_res$comparisons$level1_vs_level1$jaccard

## level2 vs. level2
paired_res$comparisons$level2_vs_level2$jaccard
```

## KS and GSEA-style comparisons

The `ks_rank`, `ks_score`, and `gsea` methods compare a retained feature set
from one signature against ranked differential-expression scores from another
signature. These methods require the compared `OmicSignature` objects to retain
their `difexp` tables. For each requested `group_label`, genes are ranked from
the largest `-log10(p_value)` in that label to the largest `-log10(p_value)` in
the contrasting label. `ks_rank` tests where the retained features fall in that
ranking, while `ks_score` compares their numeric ranking scores against the
remaining features. The legacy `method = "ks"` name is accepted as an alias for
`method = "ks_rank"`.

```{r}
ks_res <- compare_omic_signatures(
  sig_list1 = signature_list,
  method = "ks_rank",
  adj_p_cutoff = 0.05,
  min_features = 25,
  max_feature = 500
)

ks_res$comparisons$level1_vs_level1$score
ks_res$comparisons$level1_vs_level1$pvalue
```

The simulated label-pairing data can also be compared with `ks_score`. In this
toy example, the treated/up/resistant labels define the first level and the
control/down/sensitive labels define the second level.

```{r}
toy_ks_score <- compare_omic_signatures(
  sig_list1 = compare_label_pairing_example,
  method = "ks_score",
  label_pairing = list(
    signature_a = c("treated", "control"),
    signature_b = c("up", "down"),
    signature_c = c("resistant", "sensitive")
  ),
  min_features = 5
)

round(toy_ks_score$comparisons$level1_vs_level1$score, 3)

signif(toy_ks_score$comparisons$level1_vs_level1$pvalue, 3)

round(toy_ks_score$comparisons$level2_vs_level2$score, 3)

signif(toy_ks_score$comparisons$level2_vs_level2$pvalue, 3)
```

GSEA requires the optional `fgsea` package.

```{r eval=FALSE}
gsea_res <- compare_omic_signatures(
  sig_list1 = signature_list,
  method = "gsea",
  adj_p_cutoff = 0.05,
  min_features = 25,
  max_feature = 500,
  minSize = 10,
  maxSize = 500
)
```

## Plot similarity heatmaps

Use `signature_similarity_heatmap()` to visualize square self-comparison output
from `compare_omic_signatures(method = "overlap")`. The heatmap function
requires the optional `ComplexHeatmap` and `circlize` packages.

```{r}
signature_similarity_heatmap(
  overlap_res,
  measure = "jaccard",
  mode = "separate",
  triangle = "upper",
  column_names_gp = grid::gpar(fontsize = 9),
  row_names_gp = grid::gpar(fontsize = 9),
  column_names_rot = 45
)
```

With `mode="combined"`, a single triangular matrix is shown, with each entry divided
into two triangles displaying the level2 (top-right) and level1 (bottom-left) similarity.
```{r}
signature_similarity_heatmap(
  overlap_res,
  measure = "jaccard",
  mode = "combined",
  triangle = "upper",
  column_names_gp = grid::gpar(fontsize = 9),
  row_names_gp = grid::gpar(fontsize = 9),
  column_names_rot = 45
)
```

For `measure = "pvalue"`, values are plotted as `-log10(pvalue)`. Larger values
therefore indicate stronger overlap enrichment.

Rank-based KS and GSEA outputs can be visualized with `measure = "score"` or
`measure = "pvalue"`. Because these matrices are directional rather than
symmetric, `mode = "combined"` draws the full square matrix (not just one
triangle), splitting each cell into two triangles that superimpose the two
`mode = "separate"` matrices: the bottom-left triangle shows
`level1_vs_level1` and the top-right triangle shows `level2_vs_level2`.
`mode = "separate"` draws those same two matrices as separate, full
heatmaps instead.

```{r}
signature_similarity_heatmap(
  ks_res,
  measure = "score",
  mode = "combined",
  triangle = "upper",
  column_names_gp = grid::gpar(fontsize = 9),
  row_names_gp = grid::gpar(fontsize = 9),
  column_names_rot = 45
)

signature_similarity_heatmap(
  ks_res,
  measure = "score",
  mode = "separate",
  column_names_gp = grid::gpar(fontsize = 9),
  row_names_gp = grid::gpar(fontsize = 9),
  column_names_rot = 45
)
```
