---
title: "Comparing Uni-directional Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Uni-directional Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This vignette demonstrates `compare_omic_signatures()` when every signature
being compared is uni-directional: a single feature set, with no `group_label`
contrast at all. See `vignette("CompareSignatures")` for the general
bi-directional workflow, and `vignette("CompareMixedDirection")` for a list
that mixes uni- and bi-directional signatures.

## The example data

`compare_unidirectional_example` contains three simulated uni-directional
signatures sharing a 100-gene universe. `uni_x` and `uni_y` share 5 of their
10 features each; `uni_z` shares nothing with either.

```{r}
library(OmicSignature)

data(compare_unidirectional_example)

## direction_type per signature
sapply(compare_unidirectional_example, \(x) x$metadata$direction_type)
```

## Overlap: a flat comparison, with no level structure

Bi-directional signatures are compared level by level (`level1_vs_level1`,
`level2_vs_level2`, ...), since `group_label` gives each one two feature sets
to pair up. Uni-directional signatures have no such pairing to make, so when
*every* signature in both `sig_list1` and `sig_list2` is uni-directional,
`compare_omic_signatures()` returns one flat comparison instead: `comparisons`
directly holds `jaccard`, `pvalue`, and `counts`, and `label_order` is `NULL`.

```{r}
uni_res <- compare_omic_signatures(
  sig_list1 = compare_unidirectional_example,
  method = "overlap",
  min_features = 3
)

names(uni_res$comparisons)
uni_res$label_order

uni_res$comparisons$jaccard
uni_res$comparisons$counts
```

## KS and GSEA are not possible without a ranking

`ks_rank`, `ks_score`, and `gsea` need a two-group contrast to build a ranked
vector to test against, which no uni-directional signature can provide. If
none of `sig_list2` is bi-directional with a `difexp` table,
`compare_omic_signatures()` cannot construct a ranking side at all, and
errors accordingly:

```{r, error = TRUE}
compare_omic_signatures(compare_unidirectional_example, method = "ks_rank")
```
