---
title: "Comparing Mixed Uni- and Bi-directional Signatures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Mixed Uni- and Bi-directional Signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

This vignette demonstrates how `compare_omic_signatures()` handles a signature
list that mixes bi-directional signatures (a `group_label` factor with two
levels, e.g. treated vs. control) with uni-directional signatures (a single
feature set, with no `group_label` contrast at all). See
`vignette("CompareSignatures")` for the general overlap/KS/GSEA workflow, and
for how `label_pairing` resolves mismatched `group_label` level order across
bi-directional signatures.

## The example data

`compare_mixed_direction_example` extends `compare_label_pairing_example`
(three simulated bi-directional signatures, `signature_a`, `signature_b`, and
`signature_c`) with one additional uni-directional signature, `signature_d`.
`signature_d`'s features overlap strongly with `signature_a`'s `treated`
label, `signature_b`'s `up` label, and `signature_c`'s `resistant` label, and
not at all with any signature's `control`/`down`/`sensitive` label.

```{r}
library(OmicSignature)

data(compare_mixed_direction_example)

## direction_type per signature
sapply(compare_mixed_direction_example, \(x) x$metadata$direction_type)
```

## Overlap: a uni-directional signature is compared against both levels

Because `signature_d` has no `group_label` contrast, `compare_omic_signatures()`
treats its whole feature set as a single set and compares it against *both*
levels of every bi-directional signature it's compared to: the same feature
set appears, unchanged, in both the `level1_vs_level1` and `level2_vs_level2`
comparison matrices.

As in `compare_label_pairing_example`, `signature_b`'s `group_label` levels are
stored in a different order than `signature_a`'s and `signature_c`'s, so
`label_pairing` is used here too, to align matching conditions:

```{r}
mixed_res <- compare_omic_signatures(
  sig_list1 = compare_mixed_direction_example,
  method = "overlap",
  label_pairing = list(
    signature_a = c("treated", "control"),
    signature_b = c("up", "down"),
    signature_c = c("resistant", "sensitive")
  ),
  min_features = 3
)

## signature_d has no group_label levels, so label_pairing does not apply to
## it; its row is NA/NA in label_order, and it is compared unchanged against
## whichever level is being evaluated in each pass.
mixed_res$label_order

mixed_res$comparisons$level1_vs_level1$jaccard["signature_d", ]
mixed_res$comparisons$level2_vs_level2$jaccard["signature_d", ]
```

`signature_d` overlaps strongly (jaccard 0.45-0.6) with the
`treated`/`up`/`resistant` level (`level1`, after pairing) of each
bi-directional signature, and shares nothing (jaccard 0) with the
`control`/`down`/`sensitive` level (`level2`).

## KS and GSEA: a uni-directional signature can only be a geneset

`ks_rank`, `ks_score`, and `gsea` rank features using the two-group contrast in
a signature's `difexp` table, so a uni-directional signature can never serve
as the ranking side (`sig_list2`), only as a geneset (`sig_list1`).
`compare_omic_signatures()` excludes it from the ranking side with a warning,
and its column in the result matrices is entirely `NA`, while it remains fully
usable as a geneset row.

```{r}
ks_res <- compare_omic_signatures(
  sig_list1 = compare_mixed_direction_example,
  method = "ks_rank",
  label_pairing = list(
    signature_a = c("treated", "control"),
    signature_b = c("up", "down"),
    signature_c = c("resistant", "sensitive")
  ),
  min_features = 3
)

## signature_d as a ranking (column) is impossible: entirely NA.
ks_res$comparisons$level1_vs_level1$score[, "signature_d"]

## signature_d as a geneset (row) against the other signatures' rankings works.
ks_res$comparisons$level1_vs_level1$score["signature_d", ]
```
