---
title: "Create OmicSignature"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "03/31/2021"
vignette: >
  %\VignetteIndexEntry{Create OmicSignature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all(".")
library(dplyr)
```
<font color = "#000000">

## 1. Components to create an OmicSignature Object 
An `OmicSignature` object contains three parts:  
 - **metadata**, a list containing metadata fields.  
  required fields: "**signature_name**", "**organism**", "**platform**", "**direction_type**", "**phenotype**".  
 - **signature**, a dataframe.  
  required columns: "**signature_symbol**", "**signature_direction**"  
  optional column: "**signature_score**"  
  "signature_symbol" should be a subset of "symbol" in difexp (if present).  
 - **difexp** (optional), a dataframe of differential expression analysis result.  
  required columns: "**id**", "**symbol**", "**score**", "**p_value**".  

Once you have the componants above, you can create your own object:  
```
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp
)
```

Or, you can read an `OmicSignature` saved in .json format using `readJson()`. See ["here"](https://montilab.github.io/OmicSignature/articles/FunOmS.html). 

## 2. Create an OmicSignature Object Step-by-Step
The example provided below is from an experiment for Myc gene reduce in mice. 
Signatures was extracted by comparing the liver of treatment and control when 
mice is 24-month old. This is a bi-directional signature example, which contains 
up and down regulated features (genes).

### 2.1. Metadata
To be saved into a OmicSignature object, the required **metadata** fields are:  
"**signature_name**", "**organism**", "**platform**", "**direction_type**", "**phenotype**".
Additional optional fields can be added, e.g., adj_p (adjusted p-value), log fold change cutoff of 
the signature, or additional experiment descriptors, which will make the information more complete.  

One option is to create metadata list by hand:

```
metadata <- list(
  "signature_name" = Myc_reduce_mice_liver_24m,
  "organism" = "Mus Musculus",
  "sample_type" = "liver",
  "phenotype" = "Myc_reduce",
  "direction_type" = "bi-directional",
  "platform" = "GPL6246",
  "adj_p_cutoff" = 0.05,
  "score_cutoff" = 7,
  "keywords" = c("Myc", "KO", "longevity"),
  "PMID" = 25619689,
  "year" = 2015
)
```

Or use the built-in function `createMetadata()` (recommended). The function reminds 
you what attributes to include. You can also provide your own customized attributes.  
Click [here](https://montilab.github.io/OmicSignature/reference/createMetadata.html) 
to see a full list of built-in attributes.  
```{r}
metadata <- createMetadata(
  # examples of build-in attributes:
  signature_name = "Myc_reduce_mice_liver_24m", # required
  organism = "Mus Musculus", # required
  phenotype = "Myc_reduce", # required
  direction_type = "bi-directional", # required
  platform = "GPL6246", # required; must be GEO platform ID
  sample_type = "liver", # must be BRENDA ontology
  adj_p_cutoff = 0.05,
  score_cutoff = 7,
  keywords = c("Myc", "KO", "longevity"),
  PMID = 25619689,
  year = 2015,
  
  # example of cursomized attributes:
  animal_strain = "C57BL/6"
)
```

**Note**: If "sample_type" is NOT a BRENDA ontology term or "platform" is NOT a valid GEO platform accession ID, you will get warnings. See how to search for the correct term to use in ["BRENDA ontology & GEO platform ID"](https://montilab.github.io/OmicSignature/articles/BRENDA.html) section.  

#### 2.1.1 Additional info for "direction_type"
`direction_type` is one of:  
 
- "uni-directional". You only have a list of significant feature names but don't know if they are up or down regulated in the treatment group, or directional information is not applicable.  

- "bi-directional". In most cases significant features can be grouped into "up" and "down" regulated features. For example, when comparing treatment _vs._ control groups, some features will be higher ("up", or "+") and some will be lower ("down" or "-") in treatment.  

- "multi-directional". Used with multi-valued categorical phenotypes (e.g., "low" _vs._ "medium" _vs._ "high"), usually analyzed by ANOVA. In this case, the "direction" column in signature table should be the phenotype's value name (e.g., "low").  

### 2.2. Diffanal results (difexp)
A differential expression analysis matrix is **optional** but **highly recommended** if you have it.  

To be saved into a OmicSignature object, the matrix' required columns are:  
"**id**", "**symbol**", "**score**", "**p_value**".  
"id" is used as an unique identifier in case there are duplicated gene symbols. Frequently used id's include probe ID, ENSEMBL ID, or unique numbers.  

Here we show an example of how to derive the `difexp` object from the results of a differential expression analysis based on the `limma` package. Output columns include logFC, AveExpr, t score, P.Value, adj.P.Val, B score, Probe.ID, gene_symbol, and gene_name.  

```{r}
difexp <- read.table(file.path(system.file("extdata", package = "OmicSignature"), "difmatrix_Myc_mice_liver_24m_raw.txt"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)
head(difexp)
```

You can manually modify the column names to match the requirements. Alternatively, you
can use the built-in function `replaceDifexpCol()`, designed to replace some 
frequently-used alternative column names. 
```{r}
colnames(difexp) <- replaceDifexpCol(colnames(difexp))
head(difexp)
```

### 2.3. Signature 
Here we create a bi-directional signature manually from the difexp object, using the `filter()` function from the `dplyr` package. In this example, we use the `score_cutoff` and `adj_p_cutoff` previously specified
in the metadata. 

```{r}
signatures <- difexp %>%
  dplyr::filter(abs(score) > metadata$score_cutoff & adj_p < metadata$adj_p_cutoff) %>%
  dplyr::select(symbol,score) %>%
  dplyr::mutate(signature_direction=ifelse(score>0,"+","-")) %>% 
  dplyr::rename(signature_symbol="symbol", signature_score="score")
head(signatures)
```
(note: if you see numbers instead of gene symbol name in the first column, please check if the "symbol" column in your difexp matrix is "character" and not accidentally be "factor")  

To be stored into OmicSignature object, **signature** need to be a **dataframe** with column **"signature_symbol"**. Also, if the signature is "bi-directional" or "multi-directional" (specified in 
`direction_type` in `metadata` list), then the column **"signature_direction"** is 
also required. "uni-directional" type does not require this column. The optional column "signature_score" is used when feature scores are available.  

Our function `standardizeSigDF()` can help you to remove duplicate rows, empty symbols in the signature dataframe, if any.
```{r}
signatures <- standardizeSigDF(signatures)
head(signatures)
tail(signatures)
```

### 2.4. Create `OmicSignature` object
We have everything we need now.  
Use `OmicSignature$new()` to create a new OmicSignature R6 object.  
```{r}
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp
)
```

Use `print()` to see its information:
```{r}
print(OmicObj)
```

You can also ask the program to print the messages while creating the 
OmicSignature Object. By default, `print_message` is set to be `FALSE`.
```{r}
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp,
  print_message = TRUE
)
```

## 3. Create an `OmicSignature` from `difexp` and `metadata`
You can by-pass the generating signature process once you are an expert. Simply provide 
cutoffs in metadata, and `OmicSigFromDifexp()` will extract signatures from the 
`difexp` provided according to those criteria, and create the `OmicSignature` object for you.  

Remember to provide cutoffs, e.g. `adj_p_cutoff` and `score_cutoff` in `metadata`, 
and make sure your input `difexp` has those columns.  

```{r}
OmicObj1 <- OmicSigFromDifexp(difexp, metadata)
OmicObj1
```

See the top signatures: 
```{r}
head(OmicObj1$signature %>% dplyr::arrange(desc(abs(signature_score))))
```

</font>
--
