---
title: "Create OmicSignatureCollection"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "10/162024"
vignette: >
  %\VignetteIndexEntry{Create OmicSignatureCollection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all(".")
library(dplyr)
```
<font color = "#000000">

## Create an OmicSignatureCollection Object
This object contains several `OmicSignature` objects to facilitate further analysis. \

An `OmicSignatureCollection` object contains two parts: \
- **metadata** \
- **OmicSigList**, a list of OmicSignature Objects \

### 1. Metadata for the Collection

The required fields for **metadata** are: \
"**collection_name**", "**description**". \
Additional optional fields can be added. \
```{r}
ColMeta <- list(
  "collection_name" = "Example_Collection",
  "description" = "An example of signature collection",
  "organism" = "Mus Musculus",
  "author" = "me"
)
```

### 2. Prepare `OmicSignature` objects 
Create or read OmicSignature objects. \
The following signatures are pseudo results for illustration purpose. \
```{r}
OmicObj1 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "OmS_example_1.json"))
OmicObj2 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "OmS_example_2.json"))
OmicObj3 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "OmS_example_3.json"))
```

### 3. Create an `OmicSignatureCollection` object 

Use `OmicSignatureCollection$new()` and provide metadata and a list of `OmicSignature` objects: 
```{r eval=FALSE}
OmicCol <- OmicSignatureCollection$new(
  OmicSigList = list(OmicObj1, OmicObj2, OmicObj3),
  metadata = ColMeta,
  print_message = FALSE
)
```

During the creation of `OmicSignatureCollection`, all input `OmicSignature` objects will be re-created to make sure they pass all the check functions. \

</font>
--
