---
title: "Create OmicSignatureCollection"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "03/31/2021"
vignette: >
  %\VignetteIndexEntry{Create OmicSignatureCollection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all(".")
library(dplyr)
```
<font color = "#000000">

## Create an OmicSignatureCollection Object
This object contains several `OmicSignature` objects to facilitate further analysis.  

An `OmicSignatureCollection` object contains two parts:  
 - metadata  
 - OmicSigList, a list of OmicSignature Objects  

### Import OmicSignature objects
Use `read.json()` function to read `OmicSignature` objects we want to put in the collection.  
```{r}
OmicObj1 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_adipose_24m_obj.json"))
OmicObj2 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_muscle_24m_obj.json"))
OmicObj3 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_liver_24m_obj.json"))
```


### Specify metadata for the collection
The required fields for **metadata** are:  
**"collection_name", "description"**.  
We can also add additional fields.  
```{r}
ColMeta <- list(
  "collection_name" = "Myc_mice_collection",
  "description" = "A collection for Myc reduced mice 24 month",
  "organism" = "Mus Musculus",
  "author" = "me"
)
```

### Create an `OmicSignatureCollection` object.  
Use `OmicSignatureCollection$new()` and provide our metadata and a list of 
`OmicSignature` objects: 
```{r}
OmicCol <- OmicSignatureCollection$new(
  OmicSigList = list(OmicObj1, OmicObj2, OmicObj3),
  metadata = ColMeta,
  print_message = FALSE
)
```
By default, `print_message` is set to `FALSE`. You can change it to `TRUE` to 
see the messages. During the creation of OmicSignatureCollection object, all 
input `OmicSignature` objects will be re-created to make sure they are all valid.  

</font>
--
