---
title: "Create OmicSignature"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "02/06/2024"
vignette: >
  %\VignetteIndexEntry{Create OmicSignature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all(".")
library(dplyr)
```
<font color = "#000000">

## 1. Components to create an OmicSignature Object 
An `OmicSignature` object contains three parts: \

- **metadata**, a list containing metadata fields. \
  required fields: "**signature_name**", "**organism**", "**platform**", "**direction_type**", "**phenotype**". \

- **signature**, a dataframe. \
  required columns: "**signature_symbol**", "**signature_direction**" \
  optional column: "**signature_score**" \
  "signature_symbol" should be a subset of "symbol" in difexp (if present). \

- **difexp** (optional), a dataframe of differential expression analysis results. \
  required columns: "**id**", "**symbol**", "**score**", "**p_value**". \

Create the object: \
```
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp
)
```

You can also read an `OmicSignature` saved in .json format using `readJson()`. See ["here"](https://montilab.github.io/OmicSignature/articles/FunOmS.html) for more details. \

## 2. Create an `OmicSignature` Object Step-by-Step
The example provided below is from an experiment for Myc gene reduce in mice. 
Signatures were extracted by comparing the liver of treatment and control when 
mice is 24-month old. This is a bi-directional signature example, which contains 
up and down regulated features (genes). \

### 2.1. Metadata
Required **metadata** fields: \
"**signature_name**", "**organism**", "**direction_type**". \
Fields not required, but highly recommended when applicable: \
"**phenotype**", "**platform**", "**sample_type**", "**covariates**", "**score_cutoff**", "**adj_p_cutoff**", "**logfc_cutoff**". \

Option 1: Create metadata by hand (not recommended because typos can occur) \
```
metadata <- list(
  "signature_name" = Myc_reduce_mice_liver_24m,
  "organism" = "Mus Musculus",
  "sample_type" = "liver",
  "phenotype" = "Myc_reduce",
  "direction_type" = "bi-directional",
  "platform" = "GPL6246",
  "adj_p_cutoff" = 0.05,
  "score_cutoff" = 7,
  "keywords" = c("Myc", "KO", "longevity"),
  "PMID" = 25619689,
  "year" = 2015
)
```

Option 2: Use function `createMetadata()` (recommended). \
This function helps remind you of the built-in attributes. The full list of current built-in attributes is shown [here](https://montilab.github.io/OmicSignature/reference/createMetadata.html). \
You can also provide your own customized attributes. \
```{r}
metadata <- createMetadata(
  # required attributes:
  signature_name = "Myc_reduce_mice_liver_24m",
  organism = "Mus Musculus",
  direction_type = "bi-directional",

  # optional and recommended:
  phenotype = "Myc_reduce",
  covariates = "none",
  platform = "GPL6246", # must be a GEO platform ID
  sample_type = "liver", # must be BRENDA ontology

  # optional cut-off attributes.
  # specifying them can facilitate the extraction of signatures.
  # feel free to delete any that are not relevant.
  logfc_cutoff = NULL,
  p_value_cutoff = NULL,
  adj_p_cutoff = 0.05,
  score_cutoff = 7,

  # other optional built-in attributes:
  keywords = c("Myc", "KO", "longevity"),
  cutoff_description = NULL,
  author = NULL,
  PMID = 25619689,
  year = 2015,

  # example of a customized attribute:
  animal_strain = "C57BL/6"
)
```

#### 2.1.1 Instruction for "sample_type" and "platform" 
If "sample_type" is NOT a BRENDA ontology term, or "platform" is NOT a valid GEO platform accession ID, you will get warnings. See how to search for the correct term in ["BRENDA ontology & GEO platform ID"](https://montilab.github.io/OmicSignature/articles/BRENDA.html). \

#### 2.1.2 Instruction for "direction_type"
`direction_type` must be one of the following: \
 
- "uni-directional". You only have a list of significant feature names, and don't know if they are up or down regulated in the treatment group, or directional information is not applicable. An example would be "genes mutated in a disease." \

- "bi-directional". Significant features can be grouped into "up" and "down" categories. For example, when comparing treatment _vs._ control groups, some features will be higher ("up", or "+") and some will be lower ("down" or "-") in the treatment group. Similarly, when the phenotype is a continuous trait, such as age, some features will increase ("up", or "+") with age, while others will decrease ("down", or "-"). \

- "multi-directional". Used with multi-valued categorical phenotypes (e.g., "low" _vs._ "medium" _vs._ "high"), usually analyzed by ANOVA. In this case, the "direction" column in signature table should be the phenotype's category (e.g., "low"). \

### 2.2. Diffanal results (difexp)
A differential expression dataframe is **optional** but **highly recommended** if available. It facilitates downstream signature extraction. \

`difexp` is a dataframe with **required** columns: \
"**id**", "**symbol**", "**score**", "**p_value**". \
"id" is used as an unique identifier in case there are duplicated symbols. Frequently used examples including: probe ID, ENSEMBL ID, UniProt ID, unique numbers. \
"score" is usually t-test statistics or Z-score. \

Some frequently used optional column names: \
"logfc", "est", "aveexpr", "se", "robust_se", "HR", "adj_p", "gene_name", "gene_annotation". \

Here we use an example out put from the differential expression analysis using the `limma` package. 

```{r}
difexp <- read.table(file.path(system.file("extdata", package = "OmicSignature"), "difmatrix_Myc_mice_liver_24m_raw.txt"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)
head(difexp)
```

We can manually modify the column names to match the requirements. Alternatively, we can use the built-in function `replaceDifexpCol()` designed to replace some 
frequently used alternative column names. 
```{r}
colnames(difexp) <- replaceDifexpCol(colnames(difexp))
head(difexp)
```

### 2.3. Signature 
To be stored into an OmicSignature object, **signature** need to be a dataframe with column **"signature_symbol"**. Also, if the signature is "bi-directional" or "multi-directional" (specified in `direction_type` in `metadata` list), then column **"signature_direction"** is also required. \
An optional column "signature_score" is recommended when feature scores are available. \

Option 1: Extract signature from difexp. \
Here we create a bi-directional signature manually from the difexp generated above, using the `score_cutoff` and `adj_p_cutoff` previously specified in the metadata. \

```{r}
signatures <- difexp %>%
  dplyr::filter(abs(score) > metadata$score_cutoff & adj_p < metadata$adj_p_cutoff) %>%
  dplyr::select(symbol, score) %>%
  dplyr::mutate(signature_direction = ifelse(score > 0, "+", "-")) %>%
  dplyr::rename(signature_symbol = "symbol", signature_score = "score")
head(signatures)
```
(note: if you see numbers instead of gene symbol name in the first column, please check if the "symbol" column in your difexp matrix is "character" and not accidentally be "factor") \

Function `standardizeSigDF()` can help remove duplicate rows, empty symbols in the signature dataframe.
```{r}
signatures <- standardizeSigDF(signatures)
head(signatures)
tail(signatures)
```

Option 2: Manually write signature. \
For uni-directional signatures: \
```
signatures <- c("gene1", "gene2", "gene3")
```

For bi-directional signatures: \
```
signatures <- c(0.45, -3.21, 2.44)
names(signatures) <- c("gene1", "gene2", "gene3")
```
The direction will be automatically determined by the score value provided. \

### 2.4. Create the `OmicSignature` object
Use `OmicSignature$new()` to create a new OmicSignature R6 object. \
```{r}
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp
)
```

You can also ask the program to print the messages while creating the 
OmicSignature Object. By default, `print_message` is set to be `FALSE`.
```{r}
OmicObj <- OmicSignature$new(
  metadata = metadata,
  signature = signatures,
  difexp = difexp,
  print_message = TRUE
)
```

Now we can `print()` to see the information: 
```{r}
print(OmicObj)
```

And use new criteria to extract new significant features: \
(note: this does _not_ change the `signature` saved in the object)
```{r}
OmicObj$extractSignature("abs(score) > 25; adj_p < 0.001")
```

See more in ["Functionalities of OmicSignature"](https://montilab.github.io/OmicSignature/articles/FunOmS.html) section. \

## 3. Create an `OmicSignature` from `difexp` and `metadata`
You can by-pass the generating signature process once you are an expert. Simply provide cutoffs (e.g. `adj_p_cutoff` and `score_cutoff`) in the `metadata`, make sure `difexp` has those columns available, and use `OmicSigFromDifexp()` to  extract significant features and create the `OmicSignature` object. \

```{r}
OmicObj1 <- OmicSigFromDifexp(difexp, metadata)
OmicObj1
```

</font>
--
