---
title: "Functionalities of OmicSignatureCollection"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "03/31/2021"
vignette: >
  %\VignetteIndexEntry{Functionalities of OmicSignatureCollection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r message=FALSE, warning=FALSE}
devtools::load_all(".")
library(dplyr)
```
<font color = "#000000">

## Create an OmicSignatureCollection Object
This object contains several `OmicSignature` objects to facilitate further analysis.  

An `OmicSignatureCollection` object contains two parts:  
 - metadata  
 - OmicSigList, a list of OmicSignature Objects  

### First create an `OmicSignatureCollection` object
This is the same code in "Create OmicSignatureCollection" section. 
```{r}
OmicObj1 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_adipose_24m_obj.json"))
OmicObj2 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_muscle_24m_obj.json"))
OmicObj3 <- readJson(file.path(system.file("extdata", package = "OmicSignature"), "Myc_reduce_mice_liver_24m_obj.json"))
ColMeta <- list(
  "collection_name" = "Myc_mice_collection",
  "description" = "A collection for Myc reduced mice 24 month",
  "organism" = "Mus Musculus",
  "author" = "me"
)
OmicCol <- OmicSignatureCollection$new(
  OmicSigList = list(OmicObj1, OmicObj2, OmicObj3),
  metadata = ColMeta,
  print_message = FALSE
)
```

## See a summary of the metadata of the OmicSignature objects in the Collection
`$metadataSummary()` will print out the metadata fields in all _OmicSignature_ 
objects stored in the _OmicSignatureCollection_.  
When parameter "only_shared" is set to be `TRUE`, only shared metadata fields 
among all _OmicSignature_ objects will be included. Otherwise, all metadata 
fields will be included.
```{r}
OmicCol$metadataSummary(only_shared = TRUE)
```

## Extract new signatures from the `OmicSignatureCollection` object
We can extract new signatures from the `OmicSignatureCollection` object using 
new criterias.  
For example, extract all features with a absolute score > 25 and adj_p < 0.001.  
By default, the features are ranked by score.  
By default, `bind` is set to be `TRUE` to output results from all *OmicSignature* 
objects as a single dataframe.  
```{r}
OmicCol$extractSignature("abs(score) > 25 & adj_p < 0.05")
```

If `bind` is set to be `FALSE`, the output of each *OmicSignature* objects are 
provided individually as one element in a list.
```{r}
OmicCol$extractSignature("abs(score) > 25 & adj_p < 0.05", bind = F)
```

</font>
--
