---
title: "Sample Type & Platform Information"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "10/03/2024"
vignette: >
  %\VignetteIndexEntry{Sample Type & Platform Info}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<font color = "#000000">

```{r}
devtools::load_all(".")
```

We use BRENDA tissue ontology to indicate the tissue or cell-line of a signature, and GEO platform accession id for platform records.  
BRENDA tissue ontology version: **10/2021**. The OBO file was downloaded at https://www.brenda-enzymes.org/ontology.php.  
GEO platform IDs are manually downloaded at https://www.ncbi.nlm.nih.gov/geo/browse/?view=platforms. 25,827 platforms were downloaded as of 2024/02.  
If you are not sure which BRENDA tissue ontology or GEO platform accession id to use, you can search for them as shown below.  

## Search for sample type and platform 
Multiple search terms are seperated by space " ". The search is case in-sensitive. \
Set `contain_all = TRUE` to show results include all search terms. \
Set `contain_all = FALSE` to show results include any of the search terms. \

```{r}
searchSampleType("mcf cell", contain_all = TRUE)
```

```{r}
searchSampleType("MCF-10 MCF-7", contain_all = FALSE)
```

```{r}
searchPlatform("illumina hiseq 3000", organism = "Homo Sapiens", contain_all = TRUE)
```

Show only accession id:
```{r}
searchPlatform("illumina hiseq 3000", organism = "Homo Sapiens", contain_all = TRUE, accession_only = TRUE)
```

</font>
--
