---
title: "Sample Type & Platform Info"
output: rmarkdown::html_vignette
author: "Vanessa Mengze Li"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Sample Type & Platform Info}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<font color = "#000000">

```{r}
devtools::load_all(".")
```

We use BRENDA tissue ontology to indicate the tissue or cell-line of a signature. BRENDA tissue ontology version: **10/2021**. The OBO file was downloaded at https://www.brenda-enzymes.org/ontology.php. \

To ensure accurate documentation of the technologies used in the experiment, we provided a predefined list of experimental platforms. \

If you are not sure which BRENDA tissue ontology or platform name to use, you can search for them.  

## Search for sample type and platform 
Multiple search terms are seperated by space " ". The search is case in-sensitive. \
Set `contain_all = TRUE` to show results include all search terms. \
Set `contain_all = FALSE` to show results include any of the search terms. \

```{r}
OmicS_searchSampleType("mcf cell", contain_all = TRUE)
```

```{r}
OmicS_searchSampleType("MCF-10 MCF-7", contain_all = FALSE)
```

Search for a single word: 
```{r}
OmicS_searchPlatform("proteomics")
```

Search for multiple words: 
```{r}
OmicS_searchPlatform(c("transcript", "single-cell"), contain_all = TRUE)
```

If `contain_all = FALSE`, results matching any of the key words will be returned. \
```{r}
OmicS_searchPlatform(c("transcript", "single-cell"), contain_all = FALSE)
```

</font>
--
